---
output: github_document
# bibliography: vignettes/references.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# WR

The **WR** package implements various win ratio methodologies for composite endpoints of death and non-fatal events, including the (stratified) proportional win-fractions (PW) regression models (Mao and Wang, 2020 <doi:10.1111/biom.13382>), (stratified) two-sample tests with possibly recurrent nonfatal event, and sample size calculation for standard win ratio test (Mao et al., 2021 <doi:10.1111/biom.13501>).

## Installation

Install **WR** from CRAN with:

``` r
install.packages("WR")
```

You can install the development version from GitHub with:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("lmaowisc/WR")
```

## Examples

Here is a basic example.

## References
